Bayesian multi-proxy reconstruction of early Eocene latitudinal temperature gradients

### 2023-09-22 manuscript/main.qmd
units of distributions added

**Coral reefs.** Today, shallow warm-water coral reefs are limited to tropical and subtropical latitudes (\~34&deg;&nbsp;N -- 32&deg;&nbsp;S), with minimum sea surface temperature tolerances (\~18&deg;C) being the primary constraint on this distribution [@johannes1983; @kleypas1999; @yamano2001]. As coral reefs reside at the upper thermal limit of the oceans today, their maximum sea surface temperature tolerance is less well-constrained, with some studies suggesting up to 35.6&deg;C in the geological past [@jones2022reef]. Nevertheless, coral reefs have frequently been recognised as tracers of past (sub-)tropical conditions [@ziegler1984; @kiessling2001]. During the Eocene, coral communities and reefs expanded across tropical and temperate latitudes, with communities found up to palaeolatitudes of 43 &deg;&nbsp;N [@Zamagni2012]. Using a compilation of Paleocene -- early Eocene coral reefs and community localities [@Zamagni2012], we generated quantitative sea surface temperature estimates for the ECCO. To do so, we extracted localities from the compilation that are inferred to be Ilerdian (early Eocene) coral reefs, and that could be confidently assigned to the EECO. We excluded coral knobs and coral-bearing mounds which might have broader climatic limits than warm-water coral reef ecosystems. This filtering resulted in four unique coral reef localities remaining for the EECO, all of which conform to the modern latitudinal range of coral reefs (\<34&deg;&nbsp;N). Subsequently, we used statistically derived temperature limits (minimum = 21&deg;C, average = 27.6&deg;C, maximum = 29.5&deg;C) from the published literature [@kleypas1999] to define a normal probability distribution of potential temperature values for coral reef localities. This normal probability distribution was defined with a mean of 27.6 and a standard deviation of 2.125, placing 97.5% of the probability density above the minimum. As the distribution of modern corals is skewed towards warmer temperatures, this approach results in 16.5% of the probability being placed on temperatures \> 29.5&deg;C, allowing for the possibility that Eocene coral reefs were adapted to warmer conditions than present-day coral reefs.

@@

1)  *Avicennia*-only assemblages ($n=2$): the absence of Rhizophoraceae is indicative of temperatures being between 15.6&deg;C (lower temperature limt of *Avicennia*) and 20.7&deg;C (lower temperature limit of *Rhizophora*). However, a value of 22.5&deg;C is ascribed as the upper temperature limit here as *Rhizophora* is rare below this temperature. We define the *Avicennia*-only temperature distribution as a normal distribution with a mean of 19.05 and a standard deviation of 1.725, resulting in 95% of the probability density being placed within the temperature limits.

2)  *Avicennia* and Rhizophoraceae assemblages ($n=5$): the presence of both groups suggests that the locality should have a minimum temperature of 20.7&deg;C (lower temperature limit of *Rhizophora*). As the upper thermal limits of *Aviciennia* and *Rhizophora* are not well established in @Quisthoudt2012, we assign the same maximum temperature limits (29.5&deg;C) as coral reef localities, because mangroves are also widely distributed throughout tropical regions. Consequently, we define the temperature distribution for this locality as a normal distribution with a mean of 25.1 and a standard deviation of 2.2, with 95% probability density within the temperature limits.

>>>>

**Coral reefs.** Today, shallow warm-water coral reefs are limited to tropical and subtropical latitudes (\~34&deg;&nbsp;N -- 32&deg;&nbsp;S), with minimum sea surface temperature tolerances (\~18&deg;C) being the primary constraint on this distribution [@johannes1983; @kleypas1999; @yamano2001]. As coral reefs reside at the upper thermal limit of the oceans today, their maximum sea surface temperature tolerance is less well-constrained, with some studies suggesting up to 35.6&deg;C in the geological past [@jones2022reef]. Nevertheless, coral reefs have frequently been recognised as tracers of past (sub-)tropical conditions [@ziegler1984; @kiessling2001]. During the Eocene, coral communities and reefs expanded across tropical and temperate latitudes, with communities found up to palaeolatitudes of 43&deg;&nbsp;N [@Zamagni2012]. Using a compilation of Paleocene -- early Eocene coral reefs and community localities [@Zamagni2012], we generated quantitative sea surface temperature estimates for the ECCO. To do so, we extracted localities from the compilation that are inferred to be Ilerdian (early Eocene) coral reefs, and that could be confidently assigned to the EECO. We excluded coral knobs and coral-bearing mounds which might have broader climatic limits than warm-water coral reef ecosystems. This filtering resulted in four unique coral reef localities remaining for the EECO, all of which conform to the modern latitudinal range of coral reefs (\<34&deg;&nbsp;N). Subsequently, we used statistically derived temperature limits (minimum = 21&deg;C, average = 27.6&deg;C, maximum = 29.5&deg;C) from the published literature [@kleypas1999] to define a normal probability distribution of potential temperature values for coral reef localities. This normal probability distribution was defined with a mean of 27.6&deg;C and a standard deviation of 2.125&deg;C, placing 97.5% of the probability density above the minimum. As the distribution of modern corals is skewed towards warmer temperatures, this approach results in 16.5% of the probability being placed on temperatures \> 29.5&deg;C, allowing for the possibility that Eocene coral reefs were adapted to warmer conditions than present-day coral reefs.

@@

1)  *Avicennia*-only assemblages ($n=2$): the absence of Rhizophoraceae is indicative of temperatures being between 15.6&deg;C (lower temperature limt of *Avicennia*) and 20.7&deg;C (lower temperature limit of *Rhizophora*). However, a value of 22.5&deg;C is ascribed as the upper temperature limit here as *Rhizophora* is rare below this temperature. We define the *Avicennia*-only temperature distribution as a normal distribution with a mean of 19.05&deg;C and a standard deviation of 1.725&deg;C, resulting in 95% of the probability density being placed within the temperature limits.

2)  *Avicennia* and Rhizophoraceae assemblages ($n=5$): the presence of both groups suggests that the locality should have a minimum temperature of 20.7&deg;C (lower temperature limit of *Rhizophora*). As the upper thermal limits of *Aviciennia* and *Rhizophora* are not well established in @Quisthoudt2012, we assign the same maximum temperature limits (29.5&deg;C) as coral reef localities, because mangroves are also widely distributed throughout tropical regions. Consequently, we define the temperature distribution for this locality as a normal distribution with a mean of 25.1&deg;C and a standard deviation of 2.2&deg;C, with 95% probability density within the temperature limits.
